---
title: "Carbon"
output: pdf_document
date: "2024-11-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(SIBER)
library(dplyr)
library(ggplot2)
library(openxlsx)
library(RColorBrewer)
```

```{r importing}
###creating the cleaned data set

uncleaned <- read.csv("~/Desktop/Catatonk/Nexus Summer 24/FINAL Catatonk.xlsx - Cleaned.csv")

#making for easier manipulation
uncleaned <- uncleaned%>%
  mutate(ID = CUMV_Number_Sample.ID, Species = factor(Species), Year = as.numeric(Year), Decade = as.numeric(Decade), d13C = δ13C, d15N = δ15N,.keep = "none")

#creating bins of five years and adding a column denoting the bin for each data point
breaks <- seq(1930,2025, 5)
labels <- seq(1930, 2020, 5)
uncleaned$five_year_group <- cut(uncleaned$Year, breaks = breaks, labels = labels, right = F)
uncleaned$five_year_group[uncleaned$Year >= 1970 & uncleaned$Year < 1980] <- 1970
uncleaned$five_year_group <- as.factor(uncleaned$five_year_group)

```

```{r cleaning}
#finding outliers
plot(uncleaned$d13C, main = NA, ylab = expression({delta}^13*C~'permille'))
hist(uncleaned$d13C, breaks = seq(-30, -18, 0.5), main = NA, xlab = expression({delta}^13*C~'permille')) #the data looks normal enough to use z scores
z_scores <- scale(uncleaned$d13C)
outlier_index <- which(abs(z_scores) > 4) #used z score > 4 based on visual elements of the scatter plot
uncleaned[outlier_index, ] #the outliers -- there is no bias towards any factor
clean <- uncleaned[-outlier_index, ] #removing outlier from the data

```

```{r infrasctructure}
#a data frame with the average d13C and d15N values for ach species
all.avg <- clean %>%
  group_by(Species)%>%
  summarize(d13C_avg = mean(d13C), d15N_avg = mean(d15N))%>%
  arrange(d15N_avg)

#a data frame with the average d13C and d15N values for each species per decade
all.avg.decade <- clean %>%
  group_by(Species, Decade)%>%
  summarize(d13C = mean(d13C), d15N = mean(d15N))%>%
  left_join(all.avg, by = "Species")%>%
  arrange(Species, Decade)

```

Delta carbon values are influenced by primary producer in a species food chain. With the stream ecosystem species staying relatively consistent across the time period, community wide trends will be explained by outside factors.

```{r investigating}
colors <- c(brewer.pal(12, "Set3"), brewer.pal(9, "Set1"), brewer.pal(3, "Set2"))

#plotting average carbon values across time for each species
ggplot(all.avg.decade, aes(Decade, d13C, color = Species))+
  geom_point()+
  geom_path()+
  scale_color_manual(values = colors, guide = "none")+
  ylim(min(all.avg.decade$d13C), max(all.avg.decade$d13C))+
  scale_x_continuous(breaks = seq(1930, 2020, by = 10))+
  ylab(expression({delta}^13*C~'permille'))+
  ggtitle(expression("Species Average " * delta^13 * C~permille~"Over Time"))
```

Here we can see that the delta carbon values are depleting across time. Pollutants entering the stream can explain this change, as they often have more depleted isotope ratios.

```{r linear regression}
#using linear regression to analyze change over time
lm.carbon = lm(d13C ~ Decade, data = clean)
summary(lm.carbon)

#seeing if a quadratic fit is more applicable, because there seems to be a stabillization/upward trend since 1980
quad.carbon = lm(d13C ~ Decade +  I(Decade^2), data = clean)
summary(quad.carbon)

```

As predicted, the data follow a quadratic fit slightly better than linear (adjusted r-squared value 0.3349 versus 0.2981, respectively). This may be attributed to restrictions on pollutants, such as described in the Clean Water Act of 1972, or agricultural shifts in the land around Catatonk Creek.


